---
title: "BIS015 Cheatsheet"
author: "Oonagh Pretorius"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Shortcuts**
*Piping* shortcut:
Windows = Shift + Ctrl + M
Mac = Shift + Command + M

Packages to load:
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
library("skimr")
library("naniar")
```

### Loading a dataset:
CSV:
```{r}
mammals <- readr::read_csv("data/mammal_lifehistories_v2.csv")
superhero_info <- readr::read_csv("data/heroes_information.csv", na = c("", "-99", "-")) # only remove the NAs here if you know the dataset!
```

Sometimes you need to check for footers etc and remove those:
```{r message=FALSE, warning=FALSE}
ecosphere <- read_csv("data/ecs21351-sup-0003-SupplementS1.csv", skip=2) %>% 
  clean_names() %>%
  slice(1:(n() - 18)) # this removes the footer
```

Check the structure of your data & class of variables:

```{r}
glimpse()
head()
summary()
```

To simply ask if there are any NAs at all, we can use any(is.na(x)) 

```{r}
any(is.na(sleep))
```
## Are there any NA's?  
The `summary()` function in base R will tally NA's by variable, but only for categorical/factor, numeric, integer, or logical columns. What about any NA values in the character columns?
```{r}
summary(msleep)
```

```{r}
msleep %>% 
  count(conservation, sort = TRUE) #gives tally of all NAs that R can see
```
OR
```{r}
amphibio %>% 
 summarize(number_nas = sum(is.na(.)))
```


### **Replacing tings with NA**
This can be very dangerous if you don't know the data well.
```{r}
amniota_tidy <- amniota %>% 
  na_if("-999")
```

**Renaming variables**
1. Manually:
```{r}
mammals <- rename(mammals, genus="Genus", wean_mass="wean mass", max_life= "max. life", litter_size="litter size", litters_per_year="litters/year")
```
```{r}
names(mammals)
```
OR
```{r}
superhero_id <- 
  rename(superhero_info,gender = "Gender",
         eye_color = "Eye color", 
         race = "Race", 
         hair_color = "Hair color", 
         height = "Height", 
         publisher = "Publisher", 
         skin_color = "Skin color", 
         alignment = "Alignment", 
         weight = "Weight" )
superhero_id
```


2. Cleaning with janitor:
```{r}
superhero_powers <- janitor::clean_names(superhero_powers)
```

**Changing classes**
```{r}
mammals %>% mutate_if(is.character, as.factor)
glimpse(mammals) #shows that all the characters were converted to factors.
```


## **Functions**

#### **select()** for columns **filter()** for observations
```{r}
fish %>%
  select(lakeid, scalelength) %>%
  filter(lakeid == "AL")
```

#### **n_distinct**
```{r}
penguins %>%
  summarize(distinct_species = n_distinct(species),
            distinct_island = n_distinct(island),
            distinct_sex = n_distinct(sex))
```

#### **mean()**
```{r}
mean(elephants$age)
```

#### **arrange()**
```{r}
fish %>% 
  select(lakeid, length, fish_id, scalelength) %>% 
  filter(lakeid=="AL") %>% 
  arrange(fish_id) # default is ascending order. can do descending:  arrange(desc(fish_id))
```

#### **mutate()** adds new columns. can even specify
```{r}
mammals %>% 
  select(genus, species, newborn) %>%
  mutate(newborn_new = ifelse(newborn == -999.00, NA, newborn))%>%
  arrange(newborn)
```

Using the **across** function we can specify individual **columns** in which to change the observations.
across() makes it easy to apply the same transformation to multiple columns, allowing you to use select() semantics inside in "data-masking" functions like summarise() and mutate()
`across()` uses `filter()` and `select()` across multiple variables. 
```{r}
mammals %>% 
  mutate(across(c("order", "family"), tolower)) # change to lower case
```
Categorical:
```{r}
penguins %>%
  summarize(across(c(species, island, sex), n_distinct))
```

Continuous:
```{r}
penguins %>%
  group_by(sex) %>% 
  summarize(across(contains("mm"), mean, na.rm=T))
```



#### **tabyl()**
```{r}
superhero_info <- janitor::clean_names(superhero_info)
tabyl(superhero_info, alignment)
```

#### **group_by()**

```{r}
penguins %>% 
  group_by(island, species) %>% #ID two categorical variables to group by
  summarize(n_penguins=n(), .groups = 'keep') # if you don't put the .groups= "keep" it will gove you a small error. Remember for later when it becomes NB
```

#### **`summarize()`**
`summarize()` will produce summary statistics for a given variable in a data frame
Example:
```{r}
elephants %>% 
  group_by(sex) %>% 
  summarise(mean_age = mean(age), mean_height = mean(height, na.rm = TRUE))
```

Asking for min, max, mean, n in summarise():
```{r}
elephants %>% 
  filter(age > 20) %>% 
  group_by(sex) %>% 
  summarise(min_height = min(height), 
            max_height = max(height), 
            mean_height = mean(height),
            total=n())
```

Summarize across all continuous variables:
```{r}
penguins %>%
  summarise_all(mean, na.rm=T)
```

#### **Extracting specific values**
```{r}
superhero_powers %>% 
  filter(hero_names == "Angel") %>%
  select_if(all_vars(.=="TRUE"))
```

#### **count()**
`count()` is an easy way of determining how many observations you have within a column. It acts like a combination of `group_by()` and `n()`.
```{r}
penguins %>% 
  count(island, sort = T) #sort=T sorts the column in descending order
```

Tabyl has nicer output for counts
```{r}
penguins %>% 
  tabyl(island, species)
```

```{r}
penguins %>% 
  tabyl(species, island) %>% 
  adorn_percentages() %>%
  adorn_pct_formatting(digits = 2)
```

```{r}
penguins %>% 
  group_by(species) %>%
  summarise(across(c(bill_length_mm, bill_depth_mm, flipper_length_mm),
                    mean, na.rm=T), total=n())
```


#### **slice_max()** and **slice_min()** will give you only the max or min observation
```{r}
fisheries_tidy %>% 
  filter(year == "2000") %>% 
  group_by(country) %>% 
  summarize(max_catch= max(catch, na.rm = TRUE)) %>% 
  #arrange(desc(max_catch)) #we can either arrange the whole list, or just extract the max value with slice_max()
  slice_max(max_catch)
```

Filtering a range of years
```{r}
fisheries_tidy %>% 
  filter(between(year, 2008, 2012), asfis_species_name == "Cephalopoda") %>% 
  group_by(country) %>% 
  summarise(max_cephalopods = max(catch, na.rm = TRUE)) %>% 
  arrange(desc(max_cephalopods)) %>% 
  slice_max(max_cephalopods, n=5) # if you don't specify n, it will just give 1 value
```


## **`naniar`**
`naniar` is a package that is built to manage NA's. Many of the functions it performs can also be performed using tidyverse functions, but it does provide some interesting alternatives.  

`miss_var_summary` provides a nice, clean summary of NA's across the data frame.
```{r}
naniar::miss_var_summary(amniota_tidy)
```

Change values to NAs and then rerun missing variable summary.
```{r}
life_history %>% 
  na_if("-999") %>% 
  miss_var_summary()
```

### **here()**
The package here(), helps you to find your files in separate folders. Now you don't have to reset your working directory every time!
```{r}
here() #supposed to put you in the main directory
```
You just need to tell it where to find your data
```{r}
sydney_beaches <-read_csv(here("lab8", "data", "sydneybeaches.csv")) %>% janitor::clean_names()
```

### **pivot_longer()**
transforms data into long form for use by R to make figures etc.
```{r}
heartrate %>% 
pivot_longer(-patient, #patient will not pivot
             names_to = "drug", #want column names to be moved to a new column
             values_to = "heartrate" # values moved to
             )
```
 Can combine with other functions:
```{r}
expenditures %>% 
  pivot_longer(-expenditure, 
               names_to = "year", 
               values_to = "billion_dollars") %>% 
  filter(expenditure == "Medical and Health") %>% 
  arrange(-billion_dollars)
```


You can also specify a range of columns that you want to pivot.
```{r}
billboard2 <- 
  billboard %>% 
  pivot_longer(wk1:wk76, # specify a range of columns
               names_to = "week",
               values_to = "rank",
               values_drop_na = TRUE #this will drop the NA's
               )
billboard2
```


Get other examples from 8.1




### `names_sep` helps pull these apart, but we still have "exp" and "rep" to deal with.  
```{r}
qpcr_untidy %>% 
  pivot_longer(
    exp1_rep1:exp3_rep3,
    names_to= c("experiment", "replicate"),
    names_sep="_", # this can be a big problem if there is no separator
    values_to="mRNA_expression"
  )
```
### Splitting strings:
```{r}
length_data %>% 
  transform(length = str_split(length, ";")) %>%
  unnest(length)
```



### Separate()
`separate()` needs to know which column you want to split, the names of the new columns, and what to look for in terms of breaks in the data.
```{r}
heartrate3 <- heartrate2 %>% 
  separate(patient, into= c("patient", "sex"), sep = "_")
```


### Unite()
is the opposite of separate(). Its syntax is straightforward. You only need to give a new column name and then list the columns to combine with a separation character.  Give it a try below by recombining patient and sex from `heartrate3`.  
```{r}
heartrate3 %>% 
  unite(patient_sex, "patient", "sex", sep = "_")
```


## `pivot_wider()`
The opposite of `pivot_longer()`. You use `pivot_wider()` when you have an observation scattered across multiple rows. In the example below, `cases` and `population` represent variable names not observations.  

Rules:  
+ `pivot_wider`(names_from, values_from)  
+ `names_from` - Values in the `names_from` column will become new column names  
+ `values_from` - Cell values will be taken from the `values_from` column  

### `pivot_wider()` with multiple variables  
```{r}
edu_level %>% 
pivot_wider(names_from = (education_level), #new column names come from the education_level column
                values_from = c(mean_income, n)) #values come from two separate columns
```

geom_bar() for counts
geom_col() - for specifying x and y
