---
title: "Lab 9 Intro"
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("jpeg")
library("here")
library("tidyverse")
library("janitor")
```

## Seating
1. If you already have a group, it will be helpful to start sitting next to each other.
2. If you don't have a group, hang tight we are working on it.
3. Please setup your computer as normal for today.

## Warm-up
1. In the data folder there is an epidemiology data set on an outbreak of malaria.
2. `rdt` refers to rapid diagnostic test and they are identified here by age group.
3. Make the data tidy and store them as a new object.

```{r}
malaria <-read_csv(here("data", "malaria.csv")) %>% janitor::clean_names()
names(malaria)
```
```{r}
malaria2 <- malaria %>% 
  pivot_longer(-c("location_name", "data_date", "submitted_date", "province", "district", "newid", "malaria_tot"), 
               names_to = "age_group", 
               values_to = "cases",
               values_drop_na = TRUE) %>% 
              select(newid, data_date, submitted_date, location_name, province, district, age_group, cases) 
               
malaria2
```
4. Which district had the highest *total* number of cases on July 30, 2020? *Dingo district hasd the highest total number of cases.*
```{r}
malaria2 %>% 
  filter(data_date == "2020-07-30") %>% 
  group_by(district) %>% 
  summarise(tot_cases = max(cases, na.rm = TRUE)) %>% 
  arrange(-tot_cases)
```





