---
title: "Lab 6 Warm-up"
author: "O Pretorius"
date: "2023-01-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
```

### Warm-up
1. Open the data `ecs21351-sup-0003-SupplementS1.csv`
```{r}
ecosphere <- readr::read_csv("data/ecs21351-sup-0003-SupplementS1.csv")
View(ecosphere)
```
Notice strange character in some cells e.g. "?"
Multiple observations in same cell (Genus and species in same cell)

2. Clean the names of the variables
Looks weird because R tried to number the columns for you - unhelpful
You want to start with row 2
```{r}
names(ecosphere)
```

add the skip arguement to skip first two row of data
```{r}
ecosphere <- readr::read_csv("data/ecs21351-sup-0003-SupplementS1.csv", skip =2)
```

```{r}
names(ecosphere)
```
 
```{r}
ecosphere <- janitor::clean_names(ecosphere)
```
OR
```{r}
#ecosphere %>% clean_names(ecosphere)
```

3. Explore the data using a function of your choice

```{r}
glimpse(ecosphere)
```

4. Are there more veggie, omni, insecivore, or nectar feeders in the data?

```{r}
ecosphere %>% 
  tabyl(diet) %>% 
  arrange(desc(n))
```

5. Which families are Veggie?
```{r}
ecosphere %>% 
  filter(diet == "Vegetation") %>% 
  select(family, diet) #%>% 
  #tabyl(diet)
```

### tabyl can do strange things when data are converted to factors

